---
title: "Effect of imputation on rare-PCA"
output: html_notebook
---


One of the reviewers asked how practical it was to use rare variants for PCA, especially when such variants are usually imputed in large studies from SNP arrays. We tested how imputation of rare variants impacts the PCA by masking out all variants with frequency less than 0.01 and imputing them using BEAGLE. 

We used the phased sequence of 9000 individuals and imputed genotypes for an independent set of 9000 individuals. We carried out (rare) PCA using PLINK on the imputed dosage for rare variants with minor allele counts of 2, 3, or 4 in the population.

Let's plot the first two PCs.

```{r}
library(ggplot2)
library(dplyr)
library(data.table)
library(rprojroot)
library(tidyr)
library(patchwork)

F = is_rstudio_project$make_fix_file()

```


Let's look at the imputation accuaracy scores for these variants (only for chromosome 1). 

```{r}

#read file with VCF INFO column
dr2 = fread(F("gwas/grid/genotypes/tau100/ss500/train/imputed_genotypes/imputed_chr1.beagle.dr2.gz"), header=FALSE, sep="\t")

#separate info column
dr2 = dr2 %>%
  separate(V1, into = c("DR2","AF","IMP"),
           sep = ";")%>%
  separate(DR2, into = c("dr2.name","dr2.value"),
           sep = "=")%>%
  separate(AF, into = c("af.name","af.value"),
           sep = "=")

dr2$dr2.value = as.numeric(dr2$dr2.value)
dr2$af.value = as.numeric(dr2$af.value)

dr2$af.cat = cut(dr2$af.value,
                 breaks = seq(1e-05, 1e-02, 
                              by = 1e-05))

dr2.mean = dr2 %>%
  group_by(af.value) %>%
  summarize(dr2.mean = mean(dr2.value))

plt1 = ggplot(dr2%>%
         filter(af.value>0 & 
                  af.value<0.01 & 
                  IMP =="IMP"),
         aes(af.value,dr2.value),
         alpha=0.2)+
  geom_point()+
  stat_smooth()+
  theme_classic()+
  labs(x = "Allele frequency",
       y = bquote("Beagle dosage"~r^2),
       title = "A) Imputation accuracy",
       fill = "Variant\n count")+
  scale_x_log10()+
  scale_fill_viridis_c()+
  theme(legend.position = c(0.85,0.5))+
  stat_smooth()

plt1

```

Imputation accuracy decreases with allele frequency as expected. Let's look at how well the imputed variants capture population structure.

```{r}

#read rare PCs
re = fread(F("gwas/grid/genotypes/tau100/ss500/train/imputed_genotypes/imputed_chr1_20.re.all.pca.eigenvec"))

colnames(re)[1] = "FID"

#read info of which deme each individual comes from
pop100 = fread(F("gwas/grid/genotypes/tau100/ss500/iid_train.txt"))

#define bivariate color scheme for pca
d<-expand.grid(lon=1:6,lat=1:6)
d$deme<-seq(1:36)-1

r1 <- with(d,(lon-min(lon))/(max(lon)-min(lon)))
r2 <- with(d,(lat-min(lat))/(max(lat)-min(lat)))
r3 <- 1-sqrt((r1^2+r2^2)/2)
cc <- rbind(r1, r2, r3, 0.5)
rownames(cc) <- c("red", "green", "blue", "alpha")
cols <- rgb(t(cc))
d$cols=cols

#assign colors to each deme
pop100 = merge(pop100,d,by="deme")

#add this info to the eigenvectors
re = merge(re, pop100, by = c("FID","IID"))

#plot first two PCs
plt2 = ggplot(re,aes(PC1,PC2,color = as.factor(deme)))+
  geom_point(alpha = 0.5, size = 0.1)+
  theme_bw()+
  scale_color_manual(values=cols)+
  theme(legend.position="none",
        axis.text=element_text(size=10),
        panel.grid = element_blank(),
        plot.margin = unit(rep(0.5,4), "pt"),
        plot.background = element_blank(),
        plot.title = element_text(hjust = 0) )+
  labs(title = "B) PCA on imputed rare variants \n (MAC: 2-4)")

plt2


```


It's clear that the imputed genotypes, despite being potentially noisy, retain information about population structure.

Let's look at IBD-sharing PCA

```{r}


ibd.pca = fread(F("gwas/grid/genotypes/germline/germ.cm_ibd.eigenvec"))

colnames(ibd.pca) = c("FID", "IID", paste("PC", seq(1,20), sep = ""))

pop.old = fread(F("gwas/grid/genotypes/germline/grid_d36_s250_e2.pheno"))

ibd.pca = ibd.pca[, c("msp","iid") := tstrsplit(IID, split = "_")]

ibd.pca[, iid := paste("msp", as.numeric(iid) - 1, sep = "_")]

ibd.pca = merge(pop.old, ibd.pca, by.x = "IID", by.y = "iid", all.x = TRUE, sort = FALSE)

ibd.pca = merge(ibd.pca, d, by = "deme")

plt3 = ggplot(ibd.pca, aes(PC1,PC2,color = as.factor(deme)))+
  geom_point(alpha=0.5, size=0.1)+
  theme_bw()+
  scale_color_manual(values=cols)+
  theme(legend.position="none",
        axis.text=element_text(size=10),
        panel.grid = element_blank(),
        plot.margin = unit(rep(0.5,4), "pt"),
        plot.background = element_blank(),
        plot.title = element_text(hjust = 0) )+
  labs(title = "C) PCA on IBD-sharing \n (PC1 v PC2)")

plt4 = ggplot(ibd.pca, aes(PC2,PC3,color = as.factor(deme)))+
  geom_point(alpha=0.5, size=0.1)+
  theme_bw()+
  scale_color_manual(values=cols)+
  theme(legend.position="none",
        axis.text=element_text(size=10),
        panel.grid = element_blank(),
        plot.margin = unit(rep(0.5,4), "pt"),
        plot.background = element_blank(),
        plot.title = element_text(hjust = 0) )+
  labs(title = "D) PCA on IBD-sharing \n (PC2 v PC3)")



```

Let's combine these plots and save

```{r}

plt.combined = (plt1 + plt2) / (plt3 + plt4)

ggsave(F("plots/plt_imputation_rarePCA_IBDPCA.png"),
       plt.combined,
       height=7,width=7, units="in")

plt.combined

```









