---
title: "Examining prediction accuracy as a consequence of effect sizes estimated in the discovery vs an independent sample"
output: html_notebook
---


# Introduction

We were seeing that prediction accuracy, measured as the correlation between the polygenic score and genetic value, was much higher (~2x) when the variants were discovered in one sample (N = 9K) but the effects were re-estimated in siblings (N = 9k). This prediction accuracy was even higher than a fully siblig gwas (also 9k) where presumably the effects are more unbiased (not as impacted by stratification).

We think this may have something to do with winner's curse or the fact that the increase in accuracy is due to the fact that the effects are re-estimated in an independent sample. To test this, let's ignore the siblings and calculate PRS in two ways:

1. effects estimated in the discovery sample (this is what is normally done).

2. variants discovered in a GWAS in unrelated individuals and effects re-estimated in an independent sample.

Let's calculate both bias and prediction accuracy in both ways.

```{r}

library(ggplot2)
library(data.table)
library(dplyr)
library(rprojroot)
library(patchwork)

F = is_rstudio_project$make_fix_file()

```

GWAS

```{r}

#effects discovered and estimated in training set and
#prs predicted in 3rd set (used to construct sibling haplotypes)
prs1 = fread(F("gwas/grid/genotypes/tau100/ss500/revisions/prs_prediction/prs1sample/prs.smooth.a1_p3.all.nc.sscore"))

colnames(prs1) = c("rep","IID","pcs0","cm","re","cmre")

mprs1 = reshape2::melt(prs1,id.vars = c("rep","IID"),
                       value.name = "prs",
                       variable.name = "correction")

#load genetic values for individuals in the sample we are predicting - to calculate prediction accuracy
gvalue1 = fread(F("gwas/grid/genotypes/tau100/ss500/revisions/prs_prediction/gvalues/gvalue.p3.all.sscore"))

colnames(gvalue1) = c("rep","IID","gvalue")

#load latitude information - to calculate bias
pop1 = fread(F("gwas/grid/genotypes/tau100/ss500/iid_sib.txt"))

#add latitude info
mprs1 = merge(mprs1, pop1, by="IID")
#add genetic value
mprs1 = merge(mprs1, gvalue1, by=c("rep","IID"))

#center the prs and subtract out genetic value
mprs1 = mprs1%>%
  group_by(rep,correction)%>%
  mutate(prs.adjusted = prs - mean(prs),
         prs.adjusted = prs.adjusted - gvalue)

#calculate the bias and prediction accuracy 
mprs1.bias = mprs1 %>%
  group_by(rep,correction)%>%
  summarize(rlat = cor(prs,latitude),
            r2 = cor(prs,gvalue)^2)

plt_bias.all = ggplot(mprs1.bias,aes(rlat))+
  geom_histogram()+
  facet_wrap( ~ correction,
              labeller = as_labeller(c(
                pcs0 = "No correction",
                cm = "Common-PCA",
                re = "Rare-PCA",
                cmre = "Common + rare"
              )))+
  theme_classic()+
  labs(x = bquote(rho*"(polygenic score, latitude)"),
       y = "Count",
       title = "Bias in polygenic scores")

plt_bias.all

```

Rare variants more appropriately correct for stratification, that much we already knew. Let's just get the plot for "no correction", which is what we need.

```{r}

plt1.bias = ggplot(mprs1.bias%>%
                     filter(correction=="pcs0"),
                   aes(rlat))+
  geom_histogram()+
  theme_classic()+
  labs(x = bquote(rho*"(polygenic score, latitude)"),
       y = "Count",
       title = "Bias")

plt1.r2 = ggplot(mprs1.bias%>%
                   filter(correction=="pcs0"),
                 aes(r2))+
  geom_histogram()+
  theme_classic()+
  labs(x = bquote(rho^2*"(polygnenic score, genetic value)"),
       y = "Count",
       title = "Prediction accuracy")

```

Now, let's plot both bias and prediction accuracy if we ascertain variants in one sample and re-estimate in another.

```{r}

prs2 = fread(F("gwas/grid/genotypes/tau100/ss500/revisions/prs_prediction/prs/a1_r2_p3.smooth.pcs0.all.sscore"))

colnames(prs2) = c("rep","IID","pcs0","cm","re","cmre")

mprs2 = reshape2::melt(prs2,id.vars = c("rep","IID"),
                       value.name = "prs",
                       variable.name = "correction")

#load genetic values for individuals in the sample we are predicting - to calculate prediction accuracy
#same as the gvalue1

#add latitude info
mprs2 = merge(mprs2, pop1, by="IID")
#add genetic value
mprs2 = merge(mprs2, gvalue1, by=c("rep","IID"))

#center the prs and subtract out genetic value
mprs2 = mprs2%>%
  group_by(rep,correction)%>%
  mutate(prs.adjusted = prs - mean(prs),
         prs.adjusted = prs.adjusted - gvalue)

#calculate the bias and prediction accuracy 
mprs2.bias = mprs2 %>%
  group_by(rep,correction)%>%
  summarize(rlat = cor(prs,latitude),
            r2 = cor(prs,gvalue)^2)

plt_bias.all = ggplot(mprs2.bias,aes(rlat))+
  geom_histogram()+
  facet_wrap( ~ correction,
              labeller = as_labeller(c(
                pcs0 = "No correction",
                cm = "Common-PCA",
                re = "Rare-PCA",
                cmre = "Common + rare"
              )))+
  theme_classic()+
  labs(x = bquote(rho*"(polygenic score, latitude)"),
       y = "Count",
       title = "Bias in polygenic scores")

plt_bias.all

```

The bias is much smaller when the effects are re-estimated.

```{r}

plt2.bias = ggplot(mprs2.bias%>%
                     filter(correction=="pcs0"),
                   aes(rlat))+
  geom_histogram()+
  theme_classic()+
  labs(x = bquote(rho*"(polygenic score, latitude)"),
       y = "Count",
       title = "Bias")

plt2.r2 = ggplot(mprs2.bias%>%
                   filter(correction=="pcs0"),
                 aes(r2))+
  geom_histogram()+
  theme_classic()+
  labs(x = bquote(rho^2*"(polygnenic score, genetic value)"),
       y = "Count",
       title = "Prediction accuracy")



```

Let's combine these two together on the same scale.

```{r}

mprs1.bias$ascertainment = "discovery"
mprs2.bias$ascertainment = "re-estimated"

mprs.bias = rbind(mprs1.bias, mprs2.bias)
#get the xlims
xlim1 = mprs.bias%>%
  filter(correction=="pcs0")%>%
  group_by(ascertainment)%>%
  summarize(min.rlat = min(rlat),
            max.rlat = max(rlat),
            min.r2 = min(r2),
            max.r2 = max(r2))
  
 (plt1.bias + xlim(c(0.02,0.2)) + 
     geom_vline(xintercept=0,color="red",linetype="dashed")) +
 plt1.r2 + xlim(c(0.05,0.65)) + geom_vline) /
(plt2.bias + xlim(c(0.02,0.2)) +
plt2.r2 + xlim(c(0.05,0.65)))


```



```{r}

gwas1 = fread(F("gwas/grid/genotypes/tau100/ss500/train/gwas_results/fixed_effects/ge/gwas_gridt100_train.ge.1.pcs0.smooth.glm.linear.gz"))

```





